---
title: "ICEfetchR: design, decisions and open issues"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ICEfetchR: design, decisions and open issues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE)
```

ICEfetchR imports ICE futures settlement prices into the UMAR platform database
and derives position series (M01, M02, ...) from them. It currently covers one
product, ICE Low Sulphur Gasoil futures. Adding another product means adding a
row to `ice_products()` (and an expiry rule if it differs).

This document records how the package works and **why** it works that way.
Items marked *(unverified)* are working assumptions that have not been
confirmed against ICE documentation.

# 1. Data flow

```
ICE API ──► GASOIL_RAW (one series per contract)   ICE_import_structure()
                │                                   ICE_import_data_points()
                ▼                                   UMARimportR::vintage_cleanup()
            GASOIL_POS (M01..M24, derived)          ICE_import_pos_structure()
                │                                   ICE_import_positions()
                ▼                                   UMARimportR::vintage_cleanup()
            views.mat_latest_data_points            REFRESH MATERIALIZED VIEW
```

Both raw and derived series are stored as first-class platform series
(principle: *raw data is stored for transparency; heavily used derived data is
stored too, not computed in a view*). The derived series are recomputed from
the **stored** raw data, never from the in-memory scrape, so they are fully
reproducible from the database.

# 2. The ICE API

Base URL: `https://www.ice.com/marketdata/api/productguide/charting/`

| endpoint | used for | notes |
|---|---|---|
| `contract-data?productId=5817&hubId=9373&mode=2` | list of contracts (`marketId`, `marketStrip`) | `lastPrice` is the **intraday last trade**, not the settlement, and is deliberately ignored. `endDate` is the end of the *delivery month*, not the last trading day. |
| `data/historical?marketId=<id>&historicalSpan=3` | daily price history per contract | `bars` = `[date string, price]`. No volume. Today's bar only appears once the day is complete. |
| `data/current-day?marketId=<id>` | (not used) | intraday minute bars |

Quirks discovered during development:

* **The contract list varies from day to day.** We have seen 23, 28 and 30
  contracts on different days; far-dated, illiquid contracts drop out and
  reappear. Consequences: absence from the list must **not** be read as
  expiry, and a contract missing on a given day is simply updated on the next
  day it is listed (the full history is fetched each time).
* **History is (probably) a rolling ~2-year window** *(unverified)*. Every
  contract's history started on 2024-10-01, including recently listed ones.
  This is why each new vintage merges stored and fetched data (section 5).
  Other `historicalSpan` values have not been explored; a larger span could
  allow a longer one-off backfill.
* **Rate limiting.** Two quick full runs (~60 requests) triggered HTTP 429 with
  `Retry-After: 3600`. httr2's default retry then waited silently for an hour.
  `ice_request()` therefore:
    * throttles to 1 request per second (`req_throttle(rate = 1)`; the actual
      ICE limit is unknown),
    * treats 429 as **non**-transient, so it fails immediately,
    * retries only 500/502/503/504, for at most 60 seconds.
* **Request economy.** On days without new data, `prepare_ice_data()` probes
  only the 2nd and 3rd nearest contracts (all contracts share one trading
  calendar). The front contract is not used as a probe because an
  expired-but-still-listed contract would always look "not new". A no-news day
  therefore costs 3 requests instead of ~31.
* **Weekday/month parsing is locale-independent.** Bar dates such as
  `"Tue Oct 01 00:00:00 2024"` and strips such as `"Dec26"` are parsed via
  `month.abb`, never `%b`, because the production machines use a Slovenian
  locale.
* Prices look like settlement prices (they differ from `lastPrice`, and
  illiquid contracts have a price every day, as settlements do), but this is
  *(unverified)*. Only 1–3% of days are unchanged, across all maturities, so
  far contracts are not stale carry-forwards.

**Terms of use** for automated access to ice.com have not been checked.

# 3. Codes and database structure

| item | value | why |
|---|---|---|
| source | `ICE` (also for derived series) | the derived series are a mechanical remapping of ICE prices, like a rolling mean |
| tables | `GASOIL_RAW`, `GASOIL_POS` | table codes must be unique |
| raw series code | `ICE--GASOIL_RAW--202610--D` | contract as **YYYYMM** so that plain text sorting is chronological (MMYY would put `0127` before `1026`); the ICE strip (`Oct26`) is kept as `level_text` source and in messages |
| derived series code | `ICE-UMAR--GASOIL_POS--M01--D` | single hyphen in the first segment = derived by UMAR (`source.name` can never contain a hyphen, so this is unambiguous); positions zero-padded for sorting |
| non-time dimension | `contract` (raw) / `position` (derived), plus time dimension `date` | |
| dimension levels | `202610` / "October 2026"; `M01` / "Position 1 (front month)" | labels built with `month.name`, locale-independent |
| category | ICE root category, `id = 0`, for both tables | |
| unit | `usd/t` | |
| interval / period_id | `D`, ISO `YYYY-MM-DD` | business days only; gaps are handled downstream |
| `keep_vintage` | `TRUE` | unchanged history is removed by hash cleanup anyway; a revised settlement keeps the older vintage as evidence |
| `table.notes` (POS) | JSON with `derived_from`, method, expiry rule, `n_positions` | the only place where lineage of the derived table is recorded |

Product configuration lives in `ice_products()`: table code, ICE product and
hub ids, name, unit, URL, `expiry_rule`, `derived_code`, `n_positions`.

# 4. Structure import

`ICE_import_structure()` and `ICE_import_pos_structure()` are **idempotent**:
every insert is `ON CONFLICT DO NOTHING`. They run before every data import, so
newly listed contracts are registered automatically. A second run should
report 0 new rows everywhere.

`ice_get_contracts()` is called once in the runner and passed to both raw
functions via `contracts =`.

# 5. Raw data import: vintage logic

For each listed contract `prepare_series_update()`:

1. fetches the ICE history;
2. skips the contract if ICE has no history yet, and fails if ICE returns
   future dates;
3. reads the stored history of the latest vintage;
4. **skips** the contract unless ICE's last date is newer than the stored one
   (so repeated runs on the same day do nothing);
5. builds the new vintage as `merge_history(stored, fetched)`:
    * dates only in the database are kept (protects history beyond ICE's
      rolling window),
    * new dates are added,
    * where both exist, ICE wins, and changed values are counted and reported
      as revisions;
6. sets `published` to the **last data date** at UTC midnight (not the run
   time). Verified to be stored as `YYYY-MM-DD 00:00:00`.

Vintage and data inserts run inside **one transaction**
(`DBI::dbWithTransaction`). Without it, a failed data insert would leave an
empty vintage as the latest one; the next run would then see "nothing stored"
and replace the full history with ICE's 2-year window.

`UMARimportR::vintage_cleanup()` then deletes the previous vintage when the
new vintage's `partial_hash` (all values except the last) equals the old
`full_hash`. This holds whenever exactly one new day was added. After a missed
day (two new points) the old vintage is kept, which is harmless.

Why not "read, append today's price, write back"? Because the history
endpoint returns the full history, so missed runs fill themselves in and no
state has to be carried between runs.

# 6. Derived positions

## Method

Positions are **calendar-based**, not ranked:

* `front(d)` = the earliest contract month whose last trading day is on or
  after date `d`;
* `Mk(d)` = `front(d)` + (k − 1) months;
* value = the raw price of that contract on `d`, if it is stored.

Ranking the available contracts on each date would be wrong for history: the
raw table only contains contracts that were still listed when tracking began
(Oct26 onwards), so on 2024-10-02 the nearest stored contract (Oct26) was in
fact the 25th, not the front month.

`compute_positions()` maps every raw `(contract, date)` row to exactly one
position, with no loops and no per-date ranking.

## Expiry rule

`ice_expiry("GASOIL_RAW", contract_month)` = **two business days before the
14th calendar day of the delivery month**, skipping weekends only
*(rule from memory of the ICE contract specification; unverified; UK holidays
not handled)*. Examples: Oct26 → 2026-10-12, Nov26 → 2026-11-12 (the 14th is a
Saturday), Dec26 → 2026-12-10.

Safeguards:

* `compute_positions()` **fails loudly** if any raw contract has data after its
  rule-based expiry (e.g. a holiday the rule ignores pushed the real expiry
  later than computed).
* Not detectable: a holiday that makes the real expiry *earlier* than the rule.
  M01 would then be missing on that day.

The first real test is Oct26 on 2026-10-12: its last bar should carry that
date, and M01 should switch to Nov26 on 2026-10-13.

## Properties to be aware of

* **Triangular history.** M24 goes back to Oct 2024; M01 only starts on
  2026-09-11 (the day after Sep26 expired). From now on all positions fill
  completely.
* **Gaps at far positions** (M20–M24 at first) when far contracts were missing
  from the contract list on the days of import. They fill automatically once
  the contracts are listed again, because positions are recomputed in full.
  Backfilled history in the middle of a series only lands in `GASOIL_POS` on
  the next trading day (a new vintage is written only when a position has a
  newer last date).
* **Roll jumps.** On the day after an expiry each position switches to the
  next contract and drops (in backwardation) by the spread between adjacent
  contracts. The jumps are of the same size as ordinary daily moves, so they
  are invisible in charts but distort daily changes, returns and volatility.
  Day-on-day changes across roll dates are not price moves.
* **Terminology.** These are *nth-nearby* (Bloomberg: *generic*) series. They
  are **not** constant-maturity series in the strict sense (which interpolate
  to a fixed horizon and have no roll jumps). Calendar-based positions equal
  nth-nearby as long as monthly contracts are listed without gaps, which is
  true out to roughly 27 months.

## Import

`prepare_pos_data()` reads the full stored raw history
(`read_raw_history()`), recomputes all positions, and writes a new vintage only
for positions whose last date is newer than the stored one, with `published`
set to that position's **own** last date. No merge is needed: raw history only
grows. Insertion uses the same single transaction as the raw import.

# 7. Volumes

Not imported. The history endpoint has no volume, and `contract-data`'s
`volume` field is a snapshot of unclear meaning (e.g. Jan28 showed volume 801
with no last trade time) with no history. Options if needed later: log daily
`contract-data` snapshots for a couple of weeks to learn the semantics, or use
ICE Report Center end-of-day reports (access not checked).

# 8. Runner and scheduling

`ice-update-database.R` (network share, `R/`), started daily by a batch file
via Windows Task Scheduler; the log is overwritten and committed to git each
run.

```{r}
contracts <- ICEfetchR::ice_get_contracts()
ICEfetchR::ICE_import_structure("GASOIL_RAW", con = con, contracts = contracts)
raw_res <- ICEfetchR::ICE_import_data_points("GASOIL_RAW", con = con, contracts = contracts)
if (!is.null(raw_res)) {
  UMARimportR::vintage_cleanup(con, table_id = table_id("GASOIL_RAW"), schema = "platform")
  ICEfetchR::ICE_import_pos_structure("GASOIL_RAW", con = con)
  pos_res <- ICEfetchR::ICE_import_positions("GASOIL_RAW", con = con)
  if (!is.null(pos_res))
    UMARimportR::vintage_cleanup(con, table_id = table_id("GASOIL_POS"), schema = "platform")
}
# staleness check, then REFRESH MATERIALIZED VIEW only if something was inserted
```

* Cleanup, positions and the view refresh are skipped when nothing new was
  imported.
* **Staleness check**: the run fails if the latest stored raw date is more than
  6 calendar days old (covers weekends and holidays). Without it, an ICE
  outage or silent API change would produce "no new data" forever without any
  alert.
* **Email only on failure** (daily job), to the maintainer, via the shared
  `send_status_email()` pattern; table ids are looked up by code, never
  hard-coded.
* If the positions step fails after raw data went in, positions catch up on
  the next day with new raw data (full recompute), so at most about one trading
  day of lag.

# 9. Testing

* **Offline unit tests**: fetch parsing via
  `testthat::local_mocked_bindings(ice_request = ...)` with canned JSON passed
  through `jsonlite::fromJSON()`; `ice_request()` itself via
  `httr2::local_mocked_responses()` (URL, user agent, 429 fails at once);
  `merge_history()`, positions and expiry logic are pure functions.
* **dittodb fixtures** (`tests/testthat/testdata/dittodb_recording_fixtures`):
  recorded against the test database using three fixed contracts
  (`test_contracts`) and deterministic `mock_history_*()` stand-ins for ICE.
  The structure umbrella is recorded **twice**, so fixtures replay the
  "already exists" state (all counts 0). Recording and tests must use
  identical arguments (the query hash includes parameter values).
* **Umbrella functions** are tested with mocked UMARimportR and
  `DBI::dbWithTransaction` (checks: skip when nothing is new; vintages before
  data, inside one transaction), so no test writes fake prices to the test
  database.
* A live smoke test (`test-fetch-live.R`) is skipped on CI.
* Coverage: 95.85%. Remaining gaps are one-off branches (creating the ICE
  source row) and error messages.

# 10. Known limitations elsewhere in the platform

Daily data is new to the platform. Not yet adapted:

* `UMARaccessR::add_date_from_period_id()` / `add_last_date_from_period_id()`
  return `NA` for interval `D`.
* `UMARvisualisR::get_interval()` classifies `"2026-10-01"` as `"A"` (dates
  parse correctly by accident); `determine_interval()` returns `NA` for daily
  data, so growth/rolling transforms in `build_data.R` will likely fail.
* `views.mat_latest_data_points`: the `date`, `year` and `month` columns are
  `NULL` for daily data. A fix (one ISO branch, computing `date` once via
  `CROSS JOIN LATERAL`) is drafted but deferred because the view has five
  dependent views with grants. Daily users can use `period_id` directly.
* UMARaccessR "not found" conventions are inconsistent
  (`sql_get_table_id_from_table_code()` returns `NA`,
  `sql_get_source_code_from_source_name()` returns `NULL`); ICEfetchR handles
  both via `id_missing()`.
* Shelved idea: `start_date` / `end_date` columns on `platform.period`, filled
  at insert time, to replace the period-parsing code spread across packages.

# 11. TODO

## The `live` flag (before or soon after 2026-10-12)

`series.live` stays `TRUE` for expired contracts. Plan:

1. **UMARimportR**: new SQL function plus R wrapper, following the existing
   `insert_new_*` pattern:

```sql
CREATE OR REPLACE FUNCTION platform.update_series_live(p_series_id bigint,
                                                      p_live boolean,
                                                      OUT count integer)
AS $$
BEGIN
    UPDATE platform.series
       SET live = p_live
     WHERE id = p_series_id
       AND live IS DISTINCT FROM p_live;
    GET DIAGNOSTICS count = ROW_COUNT;
END;
$$ LANGUAGE plpgsql;
```

```{r}
update_series_live <- function(con, df, schema = "platform") {
  sql_function_call(con, "update_series_live",
                    list(p_series_id = df$series_id, p_live = df$live), schema)
}
```

2. **ICEfetchR**: `prepare_live_table(table_code, con, schema)` returning
   `series_id`, `live` for all raw series, with
   `live = Sys.Date() <= ice_expiry(table_code, code_to_month(contract_code))`.
   Decide from the **date rule**, never from absence in the contract list
   (which varies day to day). Position series are always live.
3. **Runner**: call it after `ICE_import_structure()`; it only changes rows
   whose status actually changes, so it is safe daily.
4. Tests: pure test of the rule; dittodb test for the prepare function.

## Checks

* 2026-10-12: Oct26's last bar is dated 2026-10-12 and M01 rolls to Nov26 on
  2026-10-13. If ICE publishes later bars, the run fails with "after
  rule-based expiry", which means the expiry rule needs fixing.
* Verify the expiry rule against the ICE contract specification, including UK
  holidays.
* Confirm the rolling history window (does `min(period_id)` move forward?) and
  try larger `historicalSpan` values for a longer backfill.
* Check ice.com terms of use for automated access.

## Later

* Daily (`D`) support in UMARaccessR and UMARvisualisR (section 10).
* `mat_latest_data_points` ISO date branch.
* Volume, if needed (section 7).
